---
title: "Environmental and Structural Drivers of Acorn Production in Coast Live Oak"
authors: "Emilin Mathew"
output: html_document
---
## Overview
Acorn production across 2024 and 2025 demonstrated minimal correlations with tree size metrics, NDVI index, and regional grouping. Box plots of acorn count across groups showed no statistically significant differences between each tree's North and South regions, with all p-values exceeding 0.1. Analysis of tree size metrics, including DBH and crown diameters, revealed weak correlations with total acorn production, with R² values below 0.07 and p-values consistently above 0.3. Similarly, 3D regression analysis of acorn production against DBH and average crown diameter in 2024 and 2025 indicated weak explanatory power (R² = 0.073 and 0.07, respectively), emphasizing the limited predictive utility of these variables. Finally, the relationship between acorn production and the NDVI across 2024, 2025, and combined years also demonstrated minimal correlation, with low R² values (0.005 to 0.045) and high p-values (p > 0.4). 


## Relationship Between NDVI and Acorn Production
This analysis examines the relationship between NDVI (Normalized Difference Vegetation Index) and acorn production across two years (2024 and 2025).

```{r setup, include=FALSE}
# Load necessary libraries
library(dplyr)
library(ggplot2)
library(lmerTest)
library(tidyr)
library(MuMIn)
library(broom.mixed)

# Set global chunk options to hide code
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)

# Load the data from CSV files
ndvi_2024 <- read.csv(here::here("static", "Tree_NDVI_Values_2024.csv"))
ndvi_2025 <- read.csv(here::here("static", "Tree_NDVI_Values_2025.csv"))
acorn_2024 <- read.csv(here::here("static", "2024_acorn.csv"))
acorn_2025 <- read.csv(here::here("static", "2025_acorn.csv"))

# Rename columns to distinguish years
data_2024 <- acorn_2024 %>%
  rename(sum_acorns_2024 = sum_acorns)

data_2025 <- acorn_2025 %>%
  rename(sum_acorns_2025 = acorn_sum)

# Merge datasets on TreeID
merged_data <- merge(data_2024, data_2025, by = "TreeID")
merged_data <- merge(merged_data, ndvi_2024, by = "TreeID")
merged_data <- merge(merged_data, ndvi_2025, by = "TreeID")

# Model 1: NDVI in 2024 predicting Acorn count in 2024
model_2024 <- lm(sum_acorns_2024 ~ NDVI_2024, data = merged_data)
summary_2024 <- summary(model_2024)
r2_2024 <- round(summary_2024$r.squared, 3)
p_value_2024 <- round(summary_2024$coefficients[2,4], 3)

# Model 2: NDVI in 2025 predicting Acorn count in 2025
model_2025 <- lm(sum_acorns_2025 ~ NDVI_2025, data = merged_data)
summary_2025 <- summary(model_2025)
r2_2025 <- round(summary_2025$r.squared, 3)
p_value_2025 <- round(summary_2025$coefficients[2,4], 3)

# Reshape data into long format for mixed model
long_data <- merged_data %>%
  pivot_longer(cols = c(sum_acorns_2024, sum_acorns_2025, NDVI_2024, NDVI_2025),
               names_to = c(".value", "Year"),
               names_pattern = "(.*)_(\\d{4})")

# Convert Year to a factor
long_data$Year <- as.factor(long_data$Year)

# Fit a mixed-effects model
model_mixed <- lmer(sum_acorns ~ NDVI + (1 | Year) + (1 | TreeID), data = long_data)

# Extract model statistics
summary_mixed <- summary(model_mixed)
r2_mixed <- round(r.squaredGLMM(model_mixed)[1], 3)  # Marginal R² (fixed effects only)
p_value_mixed <- round(coef(summary(model_mixed))["NDVI", "Pr(>|t|)"], 3)
```

### NDVI vs Acorn Production (2024)

```{r plot-2024}
ggplot(merged_data, aes(x = NDVI_2024, y = sum_acorns_2024)) +
  geom_point(color = "blue") +
  geom_smooth(method = "lm", color = "red") +
  labs(title = "NDVI vs Acorn Production (2024)", 
       x = "NDVI Index (2024)", 
       y = "Acorn Production (Caps/Quadrant, 2024)") +
  theme_minimal() +
  annotate("text", x = max(merged_data$NDVI_2024, na.rm = TRUE) * 0.95, 
           y = max(merged_data$sum_acorns_2024, na.rm = TRUE) * 0.95, 
           label = paste0("R² = ", r2_2024, "\n p = ", p_value_2024),
           hjust = 1, size = 5)
```

### NDVI vs Acorn Production (2025)

```{r plot-2025}
ggplot(merged_data, aes(x = NDVI_2025, y = sum_acorns_2025)) +
  geom_point(color = "blue") +
  geom_smooth(method = "lm", color = "red") +
  labs(title = "NDVI vs Acorn Production (2025)", 
       x = "NDVI Index (2025)", 
       y = "Acorn Production (Caps/Quadrant, 2025)") +
  theme_minimal() +
  annotate("text", x = max(merged_data$NDVI_2025, na.rm = TRUE) * 0.95, 
           y = max(merged_data$sum_acorns_2025, na.rm = TRUE) * 0.95, 
           label = paste0("R² = ", r2_2025, "\n p = ", p_value_2025),
           hjust = 1, size = 5)
```

### Combined Analysis (2024 & 2025)

```{r plot-combined}
ggplot(long_data, aes(x = NDVI, y = sum_acorns, color = Year)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "lm", se = TRUE) +
  labs(title = "NDVI vs. Acorn Production (2024 & 2025)", 
       x = "NDVI Index", 
       y = "Acorn Production (Caps/Quadrant)") +
  theme_minimal() +
  scale_color_manual(values = c("2024" = "blue", "2025" = "red")) +
  annotate("text", x = max(long_data$NDVI, na.rm = TRUE) * 0.95, 
           y = max(long_data$sum_acorns, na.rm = TRUE) * 0.95, 
           label = paste0("R² = ", r2_mixed, "\n p = ", p_value_mixed),
           hjust = 1, color = "black", size = 5)
```

### Statistical Summary

```{r stats-summary}
cat("2024 Model: R² =", r2_2024, ", p-value =", p_value_2024, "\n")
cat("2025 Model: R² =", r2_2025, ", p-value =", p_value_2025, "\n")
cat("Combined Mixed Model: R² =", r2_mixed, ", p-value =", p_value_mixed, "\n")
```